---
title: "Exercises for DES Summer Meeting workshop 2026"
author: "Benjamin Skov Kaas-Hansen"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    highlight: haddock
    theme: paper
    self_contained: true
    embed-resources: true
editor_options:
  markdown:
    wrap: 72
---

<!--
# ------------------------------------------------------------------------
#
# Consider making e.g. Shiny that produces "community-prior values" for
# both proportion and continuous variables
#
# ------------------------------------------------------------------------
-->

All of these exercises use the R program package. We will be needing a
few packages (all on CRAN). You can install these with the
`install.packages()` function:

```{r eval=FALSE}
install.packages(c(
  "rstan",
  "rstanarm",
  "bayesplot",
  "coda",
  "dplyr",
  "ggplot2",
  "purrr",
  "adaptr"
))
```

```{r message=FALSE, class.source = 'fold-show'}
for (p in c("dplyr", "ggplot2", "purrr", "adaptr", "coda")) {
  library(p, character.only = TRUE)
}
```

These exercises dive directly into real Bayesian modelling without much time
spent on "toy" examples or technicalities.

```{r echo=FALSE}
knitr::opts_chunk$set(include = TRUE, eval = TRUE)
```

## Before we start: somme intuition about sampling

```{r echo=FALSE}
knitr::opts_chunk$set(include = FALSE, eval = FALSE)
```

## A. Devising and using priors

1.  Expecting a mortality risk of 20% in patients with chronic mesenteric
    ischaemia, devise a couple of priors that express this expectation with
    multiple levels of certainty. Compute the median and 95% interval

    -   Hint: you can (but don't have to!) use `find_beta_params()` from
        the `adaptr` package

    ```{r}
    map_dfr(c(0.025, 0.05, 0.1), ~ find_beta_params(0.2, ., n_dec = 2))
    ```

2.  Devise priors with the same levels of certainty but with other
    mortality risks

    ```{r}
    expand.grid(theta = c(0.15, 0.2, 0.25), boundary = c(0.025, 0.05, 0.1)) %>%
      bind_cols(with(., map2_dfr(theta, boundary, find_beta_params, n_dec = 2)))
    ```

3.  Assume that you are working at a production facility with quality
    control of a novel CT scanner. You want to estimate the
    probability of producing a faulty machine in the coming year. In the
    previous year 365 machines were produced; none were faulty.

    Estimate the probability of faulty machine based on a frequentist
    analysis of the data. What is your estimated proportion? What is
    your forecast of faulty machines for the coming year? Can any happen?

    -   Hint: the maximum likelihood estimate of a binomial variable
        with $k$ successes (here, faulty machines) in $N$ trials (here,
        days in a year) is $\hat \theta = \tfrac{k}{N}$

    -   Hint: recall the definitions of probability from the lectures

    ```{r}
    #' The maximum likelihood estimator theta (MLE) of a binomial variable with
    #' k successes in n k/n. Because k=0, the MLE is zero, leaving no uncertainty,
    #' and so there’s no standard error, making it impossible to actually derive
    #' a confidence interval. So the forecast would be 0 failures.
    ```

4.  Use a Bayesian modeling approach to estimate the probability of a
    faulty machine for the coming year. How/why can the conclusion change?
    Which result is more reasonable?

    -   Hint: which prior is conjugate to a binomial likelihood (which
        is what we have here)?

    ```{r}
    #' For a beta(a, b) prior and a binomial(n, theta) likelihood, the posterior is
    #' beta(a + k, b + n - k), because the beta distribution is a conjugate prior
    #' to the binomial likelihood. Thus, prop_failure is the posterior
    #' distribution of the expected probability of producing a faulty machine,
    #' assuming a beta(0.5, 0.5) prior (Jeffrey’s prior).

    prob_failure <- rbeta(10000, 0.5 + 0, 0.5 + 365)
    ggplot() + geom_density(aes(prob_failure), trim = TRUE)

    ggplot() +
      geom_density(aes(rbeta(10000, 0.5 + 0, 0.5 + 365), colour = "0.5, 0.5")) +
      geom_density(aes(rbeta(10000, 1 + 0, 1 + 365), colour = "1, 1")) +
      geom_density(aes(rbeta(10000, 1 + 0, 364 + 365), colour = "1, 364"))

    # Distribution of the expected number of faults with 365 machines produced:
    ggplot() + geom_histogram(aes(prob_failure * 365), binwidth = 1)
    ```


5.  Imagine you know 10 experts with experience in the same type of
    production line as your factory. You would like to survey these
    experts and use their responses to devise an appropriate prior for
    the annual fault probability. How would you go about that?

    ```{r}
    #' One could ask them to give the number of faulty batches out of 365
    #' produced. Then, one could compute the pooled proportion and "rescale"
    #' to arrive at a 365-day time frame. For example:

    survey_responses <- c(1, 1, 0, 0, 2, 1, 2, 1, 1, 0)
    mean(survey_responses)

    ggplot() +
      stat_function(fun = ~ dbeta(., 0.9, 365 - 0.9), n = 1000, xlim = c(0, 0.05))
    ```

## B. Using rstanarm

Some claim that women are under-diagnosed, and that more men are more
likely to have CTs than women. We will try to analyse this claim using
the `rstanarm` package, using imaginary data for the CTs on a specific
day. Let's load the relevant package and create a dataset that can be
used for logistic regression.

```{r include=TRUE, message=FALSE, class.source = 'fold-show'}
library("rstanarm")
library("bayesplot")
dta <- data.frame(woman = rep(0:1, times = c(22 - 8, 8)))
```

1.  Use the `stan_glm()` function to fit a logistic model. Remember that
    a logistic model fits parameters on the log-odds (= logit) scale.

    -   Hint: remember to set `family = binomial` to run a logistic
        regression

    -   Hint: to fit a regression model without covariates, the formula should
        have the form `y ~ 1` where `y` is the outcome variable in the data.

```{r}
fit <- stan_glm(
  woman ~ 1,
  data = dta,
  family = binomial,
  cores = 1,
  refresh = 0 # just to suppress sampler output (avoids cluttering)
)
```


2.  Extract the posterior values and plot the posterior distribution of
    the intercept.

    -   Hint: use `as.matrix()` and `mcmc_areas()`

```{r}
posteriors <- as.matrix(fit)
as_tibble(fit)

mcmc_areas(posteriors)
```

3.  Compute the MAP and median posterior on the probability scale.

    -   Hint: recall that $g(x) = \frac{\exp(x)}{1 + \exp(x)}$ is the
        inverse function to the logistic transform

    -   Hint: look at `?posterior_epred`

    ```{r}
    g <- function(logit) {
      exp(logit) / (1 + exp(logit))
    }

    ggplot() +
      geom_density(aes(g(posteriors[, 1])))
    ```


4.  `stan_glm()` uses a zero-mean normal distribution as prior for the
    intercept by default. How could we modify the prior
    distribution to make it resemble a uniform distribution? Try to
    refit the model with this uniform distribution

    -   Hint: look at `?stan_glm` and
        <https://mc-stan.org/rstanarm/articles/priors.html>

    ```{r}
    fit_uniform_prior <- stan_glm(
      woman ~ 1,
      data = dta,
      family = binomial,
      prior_intercept = NULL, # uses uniform prior
      cores = 1,
      refresh = 0 # just to suppress sampler output (avoids cluttering
    )
    ```


5.  Try to change the prior distribution to e.g. `normal(0.0, 0.1)` and
    rerun the analysis. How does this impact the conclusion? Why?

    ```{r}
    fit_normal_prior <- stan_glm(
      woman ~ 1,
      data = dta,
      family = binomial,
      prior_intercept = normal(0.0, 0.1),
      cores = 1,
      refresh = 0 # just to suppress sampler output (avoids cluttering)
    )
    ```

6.  Try your own prior and see how it influences the result.

    ```{r}
    #' Same as above but using whatever prior you used
    ```

7.  Prior predictive checks are like posterior predictive checks, only that they ignore the data and generates data based solely on the priors. This can often be useful to check that the priors you have specified will yield data that look reasonable for the modelling task at hand. Carry out a prior predictive check of the three models above (default prior, `normal(0.0, 0.1)` and your own). All good?

    -   Hint: see the `prior_PD` argument in `?stan_glm`.

    ```{r}
    # Create new larger dataset
    update(fit, prior_PD = TRUE)
    ```

8.  Now assume that we had had data from 10 departments.
    Construct a new data frame akin to the one above but with more
    observations. Then, rerun the analysis with your own prior and
    compare the results from this larger dataset to those of the
    previous analysis. How are they similar? How do they differ?

    ```{r}
    # Create new larger dataset
    dta_large <- data.frame(woman = rep(0:1, times = c(220 - 80, 80)))

    fit_large <- update(fit, data = dta_large) # do the same with the other fits
    summary(fit_large)
    ```


## C. Multiple linear regression

In this exercise we will use data from a fictive department of radiolgy
with CMI patients randomised to miracustent (`treatment = 1`) vs. standard care
(`treatment = 0`) . The dataset also contains a few baseline covariates
(`age,` `woman` and `sepsis`) along with outcome variables (`icu_los`,
`hosp_los` and `dead_at_90_days`).

Download the data and take a look:

```{r include=TRUE, eval=TRUE, class.source = 'fold-show'}
cmi <- read.delim(
  "https://raw.githubusercontent.com/epiben/des-summer-meeting-2026/main/data/cmi.tsv"
)
head(cmi)
```

1.  Analyse the data with a regular frequentist model, estimating the
    effect of relevant covariates on the ICU length of stay. Ignore the
    `treatment` covariate for now. What are your overall conclusions?
    What are your conclusions about the impact of sepsis on ICU
    length-of-stay?

    -   Hint: relevant covariates should not contain information about
        the future

    ```{r}
    freq_fit <- lm(icu_los ~ age + sepsis + woman, data = cmi)
    summary(freq_fit)
    ```

2.  Analyse the same model using a Bayesian analysis with the
    `stan_glm()` function. Try to specify reasonable priors

    ```{r cache=TRUE}
    priors <- normal(
        location = c(0.5, 5, -1), scale = c(2, 2, 2)
    )
    bayes_fit <- stan_glm(
      icu_los ~ age + sepsis + woman,
      data = cmi,
      prior = priors,
      seed = 2026,
      refresh = 0,
      QR = TRUE
    )

    summary(bayes_fit)
    ```

3.  Examine the fit of the model. Convergence okay?

    ```{r}
    posterior_vs_prior(bayes_fit, pars = c("age", "sepsis1", "woman0")) +
        coord_cartesian(ylim = c(-5, 10))

    rhat(bayes_fit)
    mcmc_dens_overlay(bayes_fit)
    mcmc_trace(bayes_fit)
    ```

4.  Use shinystan to investigate the model.

    ```{r eval=FALSE}
    launch_shinystan(bayes_fit)
    ```

5.  Increase the number of chains to, say, 8 and the double the number
    of iterations. Examine the fitted model (for example with
    shinystan). What's the impact on the results? What problems could
    more and longer chains potentially fix?

    ```{r cache=TRUE}
    bayes_fit_more_longer <- stan_glm(
      icu_los ~ age + sepsis + woman,
      data = cmi,
      iter = 8000,
      chains = 8,
      seed = 2026,
      refresh = 0
    )

    summary(bayes_fit_more_longer)
    ```

6.  Look at the posterior distribution of the `sepsis` parameter. What
    is your conclusion about the effect of sepsis on ICU length of stay?
    What is the probability that the effect of sepsis is associated with
    shortened ICU-LOS? Shortening by more than 1 day?

    ```{r}
    mcmc_areas(bayes_fit, pars = "sepsis", prob = 0.95)

    mean(as.data.frame(bayes_fit)$sepsis < 0)
    mean(as.data.frame(bayes_fit)$sepsis < -1)
    mean(as.data.frame(bayes_fit)$sepsis < -2)
    mean(as.data.frame(bayes_fit)$sepsis < -5)
    ```

7.  How might you use a survey to devise an informative prior on the
    effect of age on ICU length of stay?

    ```{r}
    #' One could, for example, ask respondents to distribute 20 X's on a grid with
    #' how many days longer a patient would be expected to remain admitted for
    #' each decade they are older. Then, combine all those X's into one and fit
    #' a Normal distribution to it whose mean and sd would be the prior values.
    #'
    #' One respondents might submit this:
    #'
    #' ^
    #' |           X X X
    #' |         X X X X X
    #' |         X X X X X
    #' |       X X X X X X X
    #' |----------------------->
    ```


8.  We can use leave-one-out-cross-validation (LOO) information
    criterion (IC) to see how well different models predict the outcome.
    These LOOICs can then be compared. Like other information criteria,
    lower is better. Importantly, the LOOIC comes with a standard error.
    Fit a model without the sepsis covariate and compare the LOOIC
    between the original and the reduced models.

    -   Hint: see `?loo` and `?loo_compare`

    ```{r cache=TRUE}
    bayes_fit_wo_sepsis <- stan_glm(
      icu_los ~ age + woman,
      data = cmi,
      seed = 2026,
      refresh = 0,
      QR = TRUE  # makes things faster
    )

    loo_fit <- loo(bayes_fit)
    loo_fit_wo_sepsis <- loo(bayes_fit_wo_sepsis)
    loo_compare(loo_fit, loo_fit_wo_sepsis)
    ```


## D. Effect-size estimation in clinical trial

1.  A friend of yours wants to test a new pre-stent treatment obelixaban
    against placebo, in CMI patients with septic shock. Pre-trial
    evidence suggest that the 30-day mortality in this patient group
    receiving standard of care is roughly 20% (95% CrI: 5%-35%). Your
    friend's main results with respect to 30-day mortality are in the
    following 2x2 table (assume no loss to follow-up):

    <div>

    |                      | Placebo | Obelixaban |
    |----------------------|:-------:|:----------:|
    | No. alive at 30 days |   161   |     164    |
    | No. dead at 30 days  |    46   |      35    |

    </div>

    Using an appropriate prior, estimate the effect size of obelixaban
    compared to placebo.

    -   Hint: recall that derived quantities, based on posterior
        parameter estimates, can be computed directly with the posterior
        samples of the parameters

    -   Hint: effect-size metrics include risk ratio (RR), risk
        difference (RD) and number-needed-to-treat (NNT)

    -   Hint: use what you learnt in A1 and the example in the lecture

    ```{r}
    find_beta_params(0.2, 0.05)
    post_estimates <- tibble(
      mortality_prop_control = rbeta(10000, 3 + 46, 13 + 161),
      mortality_prop_obelixaban = rbeta(10000, 3 + 35, 13 + 164),
      RR = mortality_prop_obelixaban / mortality_prop_control
    )
    quantile(post_estimates$RR, c(0.025, 0.5, 0.975))
    ```


2.  After helping your friend with their unadjusted effect-size
    estimation above, you return to your own results of the effect of
    miracustent against standard care on 90-day mortality (same
    data as in section C). \
    Model the 90-day mortality risk using the four baseline covariates
    in the dataset (`treatment`, `age`, `woman` and `sepsis`) and
    default priors. What do you conclude about the conditional effects
    of miracustent on the 90-day mortality risk? Can you conclude
    anything about the effects of the other covariates?

    ```{r cache=TRUE}
    mortality_fit <- stan_glm(
      dead_at_90_days ~ treatment + age + woman + sepsis,
      data = cmi,
      family = binomial,
      seed = 2026,
      QR = TRUE,
      refresh = 0 # suppress fitting progress output
    )

    summary(mortality_fit, probs = c(0.025, 0.5, 0.975))

    # Convergence diagnostics
    rhat(mortality_fit)
    mcmc_dens_overlay(mortality_fit)
    mcmc_trace(mortality_fit)

    mcmc_areas(
      mortality_fit,
      point_est = "median",
      pars = vars(-"(Intercept)"),
      prob = 0.95
    ) +
      scale_x_continuous(
        breaks = log(c(0.1, 0.2, 0.33, 0.5, 0.67, 1, 1.5, 2, 3, 5, 10)),
        labels = exp
      )
    ```

3.  Since you need to get your new stent approved by the authorities, you
    decide to use a somewhat skeptical prior on the treatment effect so
    that an estimated effect will convince even the grumpiest regulator.
    How could you devise such a skeptical prior for the effect of the
    treatment?

    -   Hint: remember that coefficients in a logistic regression are
        log-OR's and that the distribution of log-OR can be expressed in
        terms of an equivalent trial of a given size with a given event
        proportion.

    ```{r}
    r <- 0.35
    N <- 200
    (sigma <- sqrt(4 /(1 - r)/N + 4/r/N))

    ggplot() +
      geom_density(aes(rlnorm(1000, meanlog = 0, sdlog = sigma))) +
      scale_x_log10(breaks = c(0.33, 0.5, 0.67, 0.8, 1, 1.25, 1.5, 2, 3))
    ```

4.  Rerun the logistic regression with this new informative prior.
    What's the impact on the (estimated) conditional treatment effect?

    ```{r message=FALSE, cache=TRUE}
    # Only difference is the scale of the first parameter
    priors <- normal(
      location = c(0, 0, 0, 0),
      scale = c(sigma, 2.5, 2.5, 2.5)
    )

    mortality_fit_skeptical <- stan_glm(
      dead_at_90_days ~ treatment + age + woman + sepsis,
      data = cmi,
      prior = priors,
      family = binomial,
      seed = 2026,
      QR = TRUE,
      refresh = 0 # suppress fitting progress output
    )

    summary(mortality_fit_skeptical, probs = c(0.025, 0.5, 0.975))

    prior_summary(mortality_fit_skeptical)

    map(list(mortality_fit, mortality_fit_skeptical), as.data.frame) %>%
      map_dfc("treatment") %>%
      setNames(c("default", "skeptical")) %>%
      ggplot() +
      geom_density(aes(x = exp(default), colour = "default")) +
      geom_density(aes(x = exp(skeptical), colour = "skeptical")) +
      scale_x_log10(breaks = c(0.1, 0.2, 0.33, 0.5, 0.67, 1, 1.5, 2, 3, 5, 10)) +
      labs(x = "OR", y = NULL)
    ```

5.  Compute average treatment effect using G-computation

    -   Hint: predict the posterior probabilities of death for all patients if
        they had all received the miracustent and if they had all received
        placebo care. Take a look at ?posterior_epred
    -   Hint: here are two links for good introductions to G-computation:
        https://github.com/kathoffman/causal-inference-visual-guides/blob/master/visual-guides/G-Computation.pdf,
        https://doi.org/10.1093%2Fije%2Fdyw323

    ```{r}
    posterior_marginal_estimates <- map_dfc(
      c(miracustent = 1, placebo = 0),
      ~ rowMeans(posterior_epred(mortality_fit, mutate(cmi, treatment = .)))
    ) %>%
      mutate(
        risk_ratio = miracustent / placebo,
        risk_diff = placebo - miracustent,
        NNT = 1/risk_diff
      )

    ggplot(posterior_marginal_estimates) +
      geom_density(aes(risk_ratio))
    ```


## E. Coding up linear regression in Stan

**NB! We didn't make it here**

Build and run your own Stan model of a linear regression as the one in
D2 above. Analyze the full data with the lung capacity as a Stan model
and run it. Argue for all the choices that you have made so far
regarding priors, likelihood, etc. Recall from the lecture that a basic
Stan program contains three sections: data, parameters, and model. Below
is the sample code for a standard linear regression analysis:

``` stan
data {
  int<lower=1> N;
  vector[N] y;
  vector[N] x;
}

parameters {
  real intercept;
  real beta;
  real<lower=0> sigma; // note lower limit
}

model {
  // Prior distributions
  intercept ~ normal(0, 10);
  beta ~ normal(0, 10);
  sigma ~ cauchy(0, 2.5);

  // Likelihood
  y ~ normal(intercept + beta * x, sigma);
}
```

1.  Modify this code to use all covariates and run the analysis

    -   Hint: start by including and using the two numeric variables
        first.

    -   Hint: then include the categorical variables as numeric
        variables (with values either 0 or 1). This works because the
        categorical predictors are binary.

## F. Resources

- Book - Richard McElreath: Statistical Rethinking [https://xcelab.net/rm/statistical-rethinking/](https://xcelab.net/rm/statistical-rethinking/). Excellent introduction to Bayesian analysis. You really should start here.
- Article - Michael Betancourt: A Conceptual Introduction to Hamiltonian Monte Carlo [https://arxiv.org/pdf/1701.02434.pdf](https://arxiv.org/pdf/1701.02434.pdf). Excellent paper on HMC; it is a bit technical, but even if you skip the most technical parts and equations, there's lots of learning in there.
- Article - Goligher et al.: Extracorporeal membrane oxygenation for severe acute respiratory distress syndrome and posterior probability of mortality benefit in a post hoc bayesian analysis of a randomized clinical trial [https://doi.org/10.1001/jama.2018.14276](10.1001/jama.2018.14276)
- Book - James V. Stone: Bayes' Rule: a tutorial introduction to Bayesian analysis. A bit old but with a fun-ish take on things and not too much math.
- Book - Andrew Gelman et al.: Bayesian Dataz Analysis. The Bible but really not accessible and in no way the place to start for most people.
